\bf Estimating the number of clusters using cross-validation

### 2018-10-02 JCGS-R2/cvclust.tex
JCGS revision 2

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}
\newcommand{\T}{T}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each
pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by
the algorithm, with light (blue) means $k = 1$ is preferred to $k = 2$, and
dark (red) indicates $k = 2$ is preferred to $k = 1$.
The black line is the theoretical boundary determined from
Proposition~\ref{prop:twoclust}.}
\label{fig:overlap-color_plot}
\end{figure}

@@

In this section, we perform a set of simulations to evaluate the performance
of our proposed method and the associated correlation correction described in
Section~\ref{sec:corr-correct}.  We compare our method with a basket of
competing methods including the Gap statistic
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, the CH-index \citep{calinski1974dendrite}, Hartigan's
statistic \citep{hartigan1975clustering}, the Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
and Bootstrap stability \citep{fang2012selection}.  We use the default
parameter settings for all competing methods.  For Gabriel cross-validation,
we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
cross-validation on the rows~($m=n/4$).  We also compare with Wold
cross-validation, which we describe in Appendix~\ref{sec:wold-cv}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting1/Facet.pdf}
\caption{Performance as correlation increases.}
\label{fig:setting1}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting2/Facet.pdf}
\caption{Performance as number of noise dimensions increases.}
\label{fig:setting2}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting3/Facet.pdf}
\caption{Performance as number of dimensions increases.}
\label{fig:setting3}
\end{figure}

Some methods, like Jump and Gap, are insensitive to higher dimensions while
other methods deteriorate quickly with increasing dimension, most notably the
Gaussian model-based BIC method.  Gabriel cross-validation and its
correlation-corrected version tend to work better in higher dimensions.

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting4/Facet.pdf}
\caption{Performance as variance heterogeneity increases.}
\label{fig:setting4}
\end{figure}

This setting demonstrates that most existing methods are sensitive to variance
heterogeneity, most notably the Gap method and the model-based BIC method. The
proposed Gabriel cross-validation method and its correlation-corrected version
consistently perform well in estimating $k$ and they are insensitive to
variance heterogeneity.

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{New simulation/demo/bench/setting5/Facet.pdf}
\caption{Performance as tail width increases.}
\label{fig:setting5}
\end{figure}

This setting investigates performance in the presence of heavy-tailed data.
When the degrees of freedom decreases, the tail becomes more flat and the
Gaussian assumption becomes more inappropriate.  For most methods, their
performances are relatively stable until the tail gets very heavy. In the case
where there are $2$ degrees of freedom, the Gap and Jump methods' performances
deteriorate considerably relative to the Gabriel method.

@@

\begin{table}
\centering
\captionsetup{justification=centering}
\caption{\label{tab:benchmark} Number of clusters selected on benchmark datasets}
\begin{tabular}{lccc}
\toprule                  
         & \multicolumn{3}{c}{Dataset} \\
\cmidrule(l){2-4}
Method   & Congress Voting & Breast Cancer & Brain Tumours \\
\midrule                    
Gabriel  & $2$             & $3$           & $5$  \\    
Gabriel (corr.~correct) & $2$ & $2$           & $5$  \\  
Wold     & $2$             & $3$           & $4$  \\
Gap      & $8$             & $10$          & $10$ \\   
BIC      & $2$         & $5$           & $2$  \\   
CH       & $2$             & $2$           & $2$  \\
Hartigan & $3$             & $3$           & $4$  \\
Jump     & $10$            & $9$           & $1$  \\   
PS       & $2$  & $2$           & $1$  \\
Stab.    & $2$  & $2$           & $7$  \\ 
\midrule 
Ground Truth & $2$         & $2$ or $3$    & $5$  \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
		\centering
	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
	\caption{Yeast data set mean expression profiles. The $5$ clusters found by
  the Gabriel method are on the bottom; Clusters profiled in
  \cite{tavazoie1999systematic} are on the right.}
	\label{fig:5_clusters}
\end{figure}

@@

\begin{figure}
\centering
\begin{minipage}{\linewidth}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{demo/elbow/correct-data.pdf}
  \end{minipage}
  \hspace{0.05in}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{demo/elbow/correct-withinss.pdf}
  \end{minipage}
\end{minipage}
\begin{minipage}{\linewidth}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{demo/elbow/incorrect-data.pdf}
  \end{minipage}
  \hspace{0.05in}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{demo/elbow/incorrect-withinss.pdf}
  \end{minipage}
\end{minipage}
\caption{Left panels show the $(X,Y)$ data points; right panels
  show the corresponding values of the within-cluster sum of squares $W_k$
plotted against the number of clusters, $k$.}
\label{fig:elbow}
\end{figure}

@@

>>>>

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\cor}{cor}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}
\newcommand{\T}{T}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{main_code/demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{main_code/demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each
pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by
the algorithm, with light (blue) means $k = 1$ is preferred to $k = 2$, and
dark (red) indicates $k = 2$ is preferred to $k = 1$.
The black line is the theoretical boundary determined from
Proposition~\ref{prop:twoclust}.}
\label{fig:overlap-color_plot}
\end{figure}

@@

In this section, we perform a set of simulations to evaluate the performance
of our proposed method and the associated correlation correction described in
Section~\ref{sec:corr-correct}.  We compare our method with a basket of
competing methods including the Gap statistic
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, the CH-index \citep{calinski1974dendrite}, Hartigan's
statistic \citep{hartigan1975clustering}, the Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Slope method \citep{fujita2014non} and Bootstrap stability \citep{fang2012selection}. 
We use the default parameter settings for all competing methods.  For Gabriel cross-validation,
we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
cross-validation on the rows~($m=n/4$).  We also compare with Wold
cross-validation, which we describe in Appendix~\ref{sec:wold-cv}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting1/Facet.pdf}
\caption{Performance as correlation increases.}
\label{fig:setting1}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting2/Facet.pdf}
\caption{Performance as number of noise dimensions increases.}
\label{fig:setting2}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting3/Facet.pdf}
\caption{Performance as number of dimensions increases.}
\label{fig:setting3}
\end{figure}

Some methods, like Jump and Gap, are insensitive to higher dimensions while
other methods deteriorate quickly with increasing dimension, most notably the
Gaussian model-based BIC method.  Gabriel cross-validation and its
correlation-corrected version tend to work better in higher dimensions.
Slope method also works better in higher dimensions.

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting4/Facet.pdf}
\caption{Performance as variance heterogeneity increases.}
\label{fig:setting4}
\end{figure}

This setting demonstrates that most existing methods are sensitive to variance
heterogeneity, most notably the Gap method, the Slope method and the model-based BIC method. The
proposed Gabriel cross-validation method and its correlation-corrected version
consistently perform well in estimating $k$ and they are insensitive to
variance heterogeneity.

@@

\begin{figure}[H]
\centering
\includegraphics[width=5.5in, height=3.3in]{main_code/demo/bench/setting5/Facet.pdf}
\caption{Performance as tail width increases.}
\label{fig:setting5}
\end{figure}

This setting investigates performance in the presence of heavy-tailed data.
When the degrees of freedom decreases, the tail becomes more flat and the
Gaussian assumption becomes more inappropriate.  For most methods, their
performances are relatively stable until the tail gets very heavy. In the case
where there are $2$ degrees of freedom, the Slope method, the Gap and Jump methods' performances
deteriorate considerably relative to the Gabriel method.

@@

\begin{table}
\centering
\captionsetup{justification=centering}
\caption{\label{tab:benchmark} Number of clusters selected on benchmark datasets}
\begin{tabular}{lccc}
\toprule                  
         & \multicolumn{3}{c}{Dataset} \\
\cmidrule(l){2-4}
Method   & Congress Voting & Breast Cancer & Brain Tumours \\
\midrule                    
Gabriel  & $2$             & $3$           & $5$  \\    
Gabriel (corr.~correct) & $2$ & $2$           & $5$  \\  
Wold     & $2$             & $3$           & $4$  \\
Gap      & $8$             & $10$          & $10$ \\   
BIC      & $2$         & $5$           & $2$  \\   
CH       & $2$             & $2$           & $2$  \\
Hartigan & $3$             & $3$           & $4$  \\
Jump     & $10$            & $9$           & $1$  \\   
PS       & $2$  & $2$           & $1$  \\
Stab.    & $2$  & $2$           & $7$  \\ 
Slope	 & $2$  & $4$           & $2$ \\
\midrule 
Ground Truth & $2$         & $2$ or $3$    & $5$  \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
		\centering
	\includegraphics[width=5.7in, height=5.7in]{main_code/data_examples/split_plots.pdf}
	\caption{Yeast data set mean expression profiles. The $5$ clusters found by
  the Gabriel method are on the bottom; Clusters profiled in
  \cite{tavazoie1999systematic} are on the right.}
	\label{fig:5_clusters}
\end{figure}

@@

\begin{figure}
\centering
\begin{minipage}{\linewidth}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{main_code/demo/elbow/correct-data.pdf}
  \end{minipage}
  \hspace{0.05in}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{main_code/demo/elbow/correct-withinss.pdf}
  \end{minipage}
\end{minipage}
\begin{minipage}{\linewidth}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{main_code/demo/elbow/incorrect-data.pdf}
  \end{minipage}
  \hspace{0.05in}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{main_code/demo/elbow/incorrect-withinss.pdf}
  \end{minipage}
\end{minipage}
\caption{Left panels show the $(X,Y)$ data points; right panels
  show the corresponding values of the within-cluster sum of squares $W_k$
plotted against the number of clusters, $k$.}
\label{fig:elbow}
\end{figure}

@@

The condition that ensures $\CV(1) < \CV(2)$ is tight but difficult to
understand intuitively. We make understand this condition better by
noting that it is equivalent to a condition on the correlation between the
two linear combinations of $X$ and $Y$ variables:
\begin{equation}\label{eqn:clust-anyd}
    \cor(u_1^\T \Sigma_{YX} X, u_1^\T Y) < 1/2.
\end{equation}
A stronger condition that is easier to understand, then, is that the canonical
correlation between $X$ and $Y$ is below $1/2$; when the latter is true the
former follows and $\CV(1) < \CV(2)$. A necessary but not sufficient
guarantee of the stronger condition is that all correlations between variables
are below $1/2$.

One might object to the condition~\eqref{eqn:clust-anyd} being stated in terms
of a particular $(X, Y)$ split of the data.  In practice, one might want the
condition to hold for all possible splits of the data. A sufficient guarantee
that~\eqref{eqn:clust-anyd} holds for any $(X, Y)$ split of the data is that
the data covariance matrix $\Sigma$ is diagonal. The ad-hoc adjustment we
propose in Sec.~\ref{sec:corr-correct} attempts to transform the data to this
form.
